---
title: "Climate Trends for Santa Cruz"
author: "Nicole Schmidt"
date: "4/5/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Attach libraries and read in data 

library(tidyverse)
library(ggplot2)
library(lubridate)
library(janitor)
library(chron)


climate <- read.table("2532221.csv", fill = TRUE,
                      sep = ',', header = T) %>% 
  clean_names()

# set date

# tidy 
climate_1900 <- climate %>% 
  filter(date >= "1900-01-01") %>% 
  filter(date < "2020-12-31") %>% 
 mutate(date = lubridate::ymd(date))

climate_1900


```

### Explore the Data

#### TMAX
```{r}
# Daily data of air temperature 

ggplot(climate_1900, aes(x = date, y = tmax))+
  geom_line()+
  labs(y = "Daily Maximum Temperature (degrees F)",
       x = "Date",
       title = "Daily Maximum Temperature in Santa Cruz from 1900-2020")+
  theme_minimal()

```

#### TMIN
```{r}

ggplot(climate_1900, aes(x = date, y = tmin))+
  geom_line()+
  labs(y = "Daily Minimum Temperature (degrees F)",
       x = "Date",
       title = "Daily Minimum Temperature in Santa Cruz from 1900 - 2020")+
  theme_minimal()


```

### Precipitation 
```{r}
# Temperature: Average of the previous and following date
# Precipitation: For dry places like the Sierra,assume no rainwhere data is missing (This maynot work for other places. Be careful!)

# Find the rows with missing data
fillrow = which(is.na(climate_1900$prcp))
fillrow
climate_1900$prcp[fillrow]=0
# replot to make sure it works 

ggplot(climate_1900, aes(x = date, y = prcp))+
geom_line()+
labs(y="Daily rainfall (inches)", x ="Date",
     title = "Daily precipitation measured in Santa Cruz from 1900 - 2021")


```

####NA for Temperature 
```{r}

fillrow = which(is.na(climate_1900$tmax) | climate_1900$tmax < 40)
fillrow = fillrow[2:length(fillrow)]
climate_1900$tmax[fillrow]=(climate_1900$tmax[fillrow+1]+climate_1900$tmax[fillrow-1])/2


ggplot(climate_1900, aes(date, climate_1900$tmax))+
  geom_line()+ labs(y="Daily Maximum Temperature (degrees f)",
                    x="Date",
                      title = "Daily maximum temperature in Santa Cruz from 1900 - 2020")+
  theme_minimal()

```

#### NA for minimum temperature
```{r}

fillrow = which(is.na(climate_1900$tmin) | climate_1900$tmin > 15)
fillrow = fillrow[2:length(fillrow)]
climate_1900$tmax[fillrow]=(climate_1900$tmax[fillrow+1]+climate_1900$tmax[fillrow-1])/2


ggplot(climate_1900, aes(date, climate_1900$tmin))+
  geom_line()+ labs(y="Daily Maximum Temperature (degrees f)",
                    x="Date",
                      title = "Daily minimim temperature in Santa Cruz from 1900 - 2020")+
  theme_minimal()


```


### Is there a trend ?
```{r}

climate_trend <- climate_1900 %>% 
  group_by(year(date)) %>% 
  summarise(tmax = mean(tmax),
            tmin = mean(tmin),
            prcp = sum(prcp))

climate_trend$dt = unique(year(date))

climate_trend

```

```{r}

ggplot(climate_trend, aes (x = dt, y = tmax))+
   labs(x = "Year", y = "Average temperature",
       title = "Trend in annual average maximum & minimum temperatures in Santa Cruz",
       subtitle = "1900 - 2020")+
  geom_point(col = "red")+
  scale_y_continuous(limits = c(climate_trend$tmin),
                     max(climate_trend$tmax))+
  geom_point(data = climate_trend, aes(x = dt, tmin), col = "blue")+
  stat_smooth(method="lm", col="red")+
  stat_smooth(data=climate_trend, aes(x=dt,tmin), col="blue", method="lm")+
  theme_minimal()

```

Now let’s calculate the slope (or how quickly temperatures are rising; we do this with linear regression)
```{r}

res=lm(tmin~dt, data=climate_trend)
summary(res)

```

```{r}
confint(res,"dt", level=0.95)
```

```{r}
ggplot(climate_trend, aes(x=dt, y=tmin))+ 
  labs(x = "Year", y = "Average minimum temperature (degrees f)",
       title = " Trend in average annual minimum temperature in Santa Cruz (1900 - 2020)")+
  stat_summary(fun.y="mean", geom="point", col="red", size=4)+
  theme(axis.text=element_text(size=14, face="bold"), 
        axis.title=element_text(size=14, face="bold"))+ 
  geom_smooth(method="lm")
```

```{r}
res=lm(tmax~dt, data=climate_trend)
summary(res)
```

```{r}
confint(res,"dt", level=0.95)
```

```{r}
ggplot(climate_trend, aes(x=dt, y=tmax)) + 
  stat_summary(fun.y="mean", geom="point", col="red", size=4)+
  theme(axis.text=element_text(size=14, face="bold"),
        axis.title=element_text(size=14, face="bold")) +
  geom_smooth(method="lm")
```

We might also cut the data into specific periods and see how the slope is changing as a function of time.
```{r}
#tmin first hald a century 

# early portion
res_late=lm(prcp~dt, 
             data=subset(climate_trend, climate_trend$dt %in% c(2000:2020)))
summary(res_early)

```

```{r}
confint(res_late,"dt", level=0.95)

```

```{r}
ggplot(subset(climate_trend, climate_trend$dt %in% c(1952:1970)), 
       aes(x=dt, y=tmin))+
  stat_summary(fun.y="mean", geom="point", col="red", size=4)+
  theme(axis.text=element_text(size=14, face="bold"), 
        axis.title=element_text(size=14, face="bold")) + 
  geom_smooth(method="lm")
```

```{r}

res_late=lm(tmin~dt, 
            data=subset(climate_trend, climate_trend$dt %in% c(1995:2021)))
summary(res_late)

```

```{r}
confint(res_late,"dt", level=0.90)
```

```{r}
# tmin lastr decades 
ggplot(subset(climate_trend,
              climate_trend$dt %in% c(1995:2019)),
       aes(x=dt, y=tmin)) + 
  stat_summary(fun.y="mean", geom="point", col="red", size=4)+
  theme(axis.text=element_text(size=14, face="bold"), 
        axis.title=element_text(size=14, face="bold")) + 
  geom_smooth(method="lm")


```

Regression assumes a linear relationship - and normally distributed data - sometimes that isn’t true, we can use non-parameteric tests to look for trends. In these cases, the Mann-Kendall test is commonly used.

tau ranges from -1 to 1 and denotes the “strength” of the trend; p-value denotes significance. Strength however can not be interpreted as slope!

```{r}

library(Kendall)
MannKendall(climate_trend$tmin)

```

```{r}
MannKendall(climate_trend$tmax)
```

```{r}

MannKendall(climate_trend$prcp)

```

We might also look at difference in means (or variance) between the two periods …Using

T-test or Rank-Sum if we think data is not normally distributed
```{r}
t.test(subset(climate_trend$tmin, climate_trend$dt %in% 1953:1969), 
       subset(climate_trend$tmin, climate_trend$dt %in% 1995:2004))
```

An alternative approach to aggregation (mean by year) is to look at a particular season, lets say we want to look only at summer (July and August)

```{r}
# create a variable
climate_1900$season = ifelse(month(date) %in% c(12,1,2), 1, ifelse(month(date) %in% c(3:5),2, ifelse(month(date) %in% c(6:8),3,4)))
clim.byseason = climate_1900 %>% group_by(year(date),season) %>% summarize(tmax=mean(tmax), tmin=mean(tmin), precip=sum(prcp))

# look only at summer
clim.summer = subset(clim.byseason, clim.byseason$season==3)
tmp=unique(year(date))
clim.summer$wy = tmp[1:length(tmp)-1]

ggplot(clim.summer, aes(x=wy, y=tmin)) + stat_summary(fun.y="mean", geom="point", col="red", size=4)+theme(axis.text=element_text(size=14, face="bold"), axis.title=element_text(size=14, face="bold")) + geom_smooth(method="lm")+labs(y=" Summer Minimum Daily Temperature C")
```

```{r}

res=lm(tmax~wy, data=clim.summer)
summary(res)

```

```{r}
confint(res,"wy", level=0.95)
```


Summer Maximum Daily Temperature 
```{r}

ggplot(clim.summer, aes(x=wy, y=tmax)) + stat_summary(fun.y="mean", geom="point", col="red", size=4)+theme(axis.text=element_text(size=14, face="bold"), axis.title=element_text(size=14, face="bold")) + geom_smooth(method="lm")+labs(y=" Summer Maximum Daily Temperature")
```

```{r}

ggplot(climate_trend, aes(x=dt, y=tmax))+ 
  stat_summary(fun.y="mean", geom="point", col="red", size=4)+
  theme(axis.text=element_text(size=14, face="bold"), 
        axis.title=element_text(size=14, face="bold"))+ 
  geom_smooth(method="lm")
```

##### TMAX from 2000 - 2020 rising significantly 
```{r}

ggplot(subset(climate_trend, climate_trend$dt %in% c(2000:2020)), 
       aes(x=dt, y=tmax))+
  labs(x = "Year", y = "Average maximum temperature (degrees f)",
       title ="Trend in average maximum temperature in Santa Cruz (2000 - 2020)")+
  stat_summary(fun.y="mean", geom="point", col="red", size=4)+
  theme(axis.text=element_text(size=14, face="bold"), 
        axis.title=element_text(size=14, face="bold")) + 
  geom_smooth(method="lm")+
  theme_minimal()
  
```
#### Trend in tmax from 1950 - 2000 to play around and change
```{r}


ggplot(subset(climate_trend, climate_trend$dt %in% c(2000:2020)), 
       aes(x=dt, y=tmin))+
  labs(x = "Year", y = "Average minimum temperature (degrees f)",
       title ="Trend in average minimum temperature in Santa Cruz (2000-2020)")+
  stat_summary(fun.y="mean", geom="point", col="red", size=4)+
  theme(axis.text=element_text(size=14, face="bold"), 
        axis.title=element_text(size=14, face="bold")) + 
  geom_smooth(method="lm")+
  theme_minimal()

```

```{r}
ggplot(climate_trend, aes (x = dt, y = prcp))+
   labs(x = "Year", y = "Average precipitation",
       title = "Trend in annual average precipitation in Santa Cruz",
       subtitle = "1900 - 2020")+
  geom_point(col = "blue")+
  stat_smooth(data=climate_trend, aes(x=dt, prcp), col="blue", method="lm")+
  theme_minimal()
```
Subset of precipitation 
```{r}
ggplot(subset(climate_trend, climate_trend$dt %in% c(1995:2020)), 
       aes(x= dt, y= prcp))+
  labs(y = "Annual precipitation (inches)", x = "Year",
       title = "Annual precipitation in Santa Cruz",
       subtitle = "1995 - 2020")+
  stat_summary(fun.y="mean", geom="point", col="blue", size=4)+
  theme(axis.text=element_text(size=14, face="bold"), 
        axis.title=element_text(size=14, face="bold")) + 
  geom_smooth(method="lm")+
  theme_minimal()
```

